A formal treatment of bidirectional typing

### 2024-01-12 tex/ESOP24/S4-pre-synthesis.tex
[ tex ] Remove \/ (LNCS has enough spacing)

\begin{theorem}\label{lem:soundness-completeness}
$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if both\/ $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and\/ $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
\end{theorem}

@@

\begin{lemma}\label{thm:Pre?-true}
If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
\end{lemma}

@@

\begin{lemma}\label{thm:Pre?-false}
If\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V |/-_{\Sigma,\Omega} t^\dir{d}$.
\end{lemma}

@@

\begin{theorem}[Generalised mode decoration]\label{thm:generalised-mode-decoration}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{theorem}

@@

\begin{lemma}\label{thm:adjustment}
For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
\begin{equation}\label{eq:strict-derivation}
V |-_{\Sigma,\Omega} t^{\dir{d'}\,g'\,\BTrue} \quad\text{for some mode~$\dir{d'}$ and boolean~$g'$}
\end{equation}
then for any mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{lemma}

@@

\begin{proposition}[Annotatability]\label{thm:annotatability}
If\/ $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that\/ $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}
Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotatecolon A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
      {(\isTerm{t} \annotatecolon A) \sqsupseteq (\isTerm{u} \annotatecolon A)}\;\Rule{Anno}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

>>>>

\begin{theorem}\label{lem:soundness-completeness}
$\Gamma |-_{\Sigma, \Omega} t :^\dir{d} A$ if and only if both $\erase\Gamma |-_{\Sigma, \Omega} t^\dir{d}$ and $\Gamma |-_{\Sigma, \erase\Omega} t : A$.
\end{theorem}

@@

\begin{lemma}\label{thm:Pre?-true}
If $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BTrue\,s}$, then $V |-_{\Sigma,\Omega} t^\dir{d}$.
\end{lemma}

@@

\begin{lemma}\label{thm:Pre?-false}
If $V |-_{\Sigma,\Omega} t^{\dir{d}\,\BFalse\,s}$, then $V |/-_{\Sigma,\Omega} t^\dir{d}$.
\end{lemma}

@@

\begin{theorem}[Generalised mode decoration]\label{thm:generalised-mode-decoration}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{theorem}

@@

\begin{lemma}\label{thm:adjustment}
For any raw term $V |-_{\Sigma,\erase\Omega} t$, if
\begin{equation}\label{eq:strict-derivation}
V |-_{\Sigma,\Omega} t^{\dir{d'}\,g'\,\BTrue} \quad\text{for some mode~$\dir{d'}$ and boolean~$g'$}
\end{equation}
then for any mode~$\dir{d}$, there is a derivation of $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some $g$~and~$s$.
\end{lemma}

@@

\begin{proposition}[Annotatability]\label{thm:annotatability}
If $\Gamma |-_{\Sigma,\erase\Omega} t : A$, then there exists~$t'$ such that $t' \sqsupseteq t$ and $\Gamma |-_{\Sigma,\Omega} t' :^\dir{d} A$ for some~$\dir{d}$.
\end{proposition}
Defined in \cref{fig:annotation-ordering}, the `annotation ordering' $t' \sqsupseteq t$ means that $t'$~has the same or more annotations than~$t$.
In a sense, annotatability is a reasonable form of completeness: if a term of a simply typed language $(\Sigma, \erase\Omega)$ is typable in the ordinary type system, it may not be directly typable in the bidirectional type system $(\Sigma, \Omega)$ due to some missing annotations, but will be if those annotations are added correctly.
In our theory, \cref{thm:annotatability} can be straightforwardly proved by induction on the derivation produced by generalised mode decoration~(\cref{thm:generalised-mode-decoration}) to construct a bidirectional typing derivation in the same mode.
The interesting case is $\SynRule{Missing}$, which is mapped to $\SynRule{Anno}$, adding to the term a type annotation that comes from the given typing derivation.
\begin{figure}
  \centering\small
  \judgbox{\isTerm{t} \sqsupseteq \isTerm{u}}{A raw term $t$ is more annotated than $u$}
  \begin{mathpar}
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
              {(\isTerm{t} \annotatecolon A) \sqsupseteq \isTerm{u}}\;\Rule{More}
    \and
    \inferrule{\vphantom{x : \Identifier}}{\isTerm{x} \sqsupseteq \isTerm{x}}\;\Rule{Var}
    \and
    \inferrule{\isTerm{t} \sqsupseteq \isTerm{u}}
      {(\isTerm{t} \annotatecolon A) \sqsupseteq (\isTerm{u} \annotatecolon A)}\;\Rule{Anno}
    \and
    \inferrule{\isTerm{t_1} \sqsupseteq \isTerm{u_1} \quad \cdots \quad \isTerm{t_n} \sqsupseteq{u_n}}
      {\tmOpts \sqsupseteq \tmOpus}\;\Rule{Op}
  \end{mathpar}
  \caption{Annotation ordering between raw terms}
  \label{fig:annotation-ordering}
\end{figure}

### 2024-01-12 tex/ESOP24/S5-synthesis.tex
[ tex ] Remove \/ (LNCS has enough spacing)

Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction on the typing derivations.
However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
\begin{lemma}[Uniqueness of synthesised types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotatecolon A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true.
        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
      \end{enumerate}
  \end{itemize}\qed
\end{proof}

@@

\begin{theorem}\label{thm:bidirectional-type-synthesis-checking}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  \begin{enumerate}
    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ for some $A$;
    \item if\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
  \end{enumerate}
\end{theorem}

@@

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we may confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$\ifarxiv (see \Cref{sec:formalisation} for details)\fi.
For the sake of clarity in presentation, though, we just use named variables and assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{defn}
By an \emph{extension}\/ $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every\/ $x$ in $\dom(\rho)$.
  By a \emph{minimal extension}\/ $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{defn}
\begin{lemma}\label{lem:unify}
  For any\/ $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution\/ $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, either
  \begin{enumerate}
    \item there is a minimal extension\/ $\bar{\rho}$ of\/ $\rho$ such that\/ $\simsub{A}{\bar{\rho}} = B$, or
    \item there is no extension\/ $\sigma$ of\/ $\rho$ such that\/ $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:

\begin{proof}[of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and independent of mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, so we only detail the last case \Rule{Op} which is new and has been discussed informally above.
  For brevity we omit the subscript $(\Sigma, \Omega)$.

>>>>

Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction on the typing derivations.
However, since mode-correctness is inductively defined on the argument list, our proof proceeds by induction on both the typing derivations and the argument list:
\begin{lemma}[Uniqueness of synthesised types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotatecolon A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true.
        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$.
      \end{enumerate}
  \end{itemize}\qed
\end{proof}

@@

\begin{theorem}\label{thm:bidirectional-type-synthesis-checking}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  \begin{enumerate}
    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ for some $A$;
    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any~$A$ whether $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A$.
  \end{enumerate}
\end{theorem}

@@

\begin{remark}
To make the argument above sound, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set~$\Xi$ of type variables has a decidable equality, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we may confine $\Xi$ to any set within the family of sets $\Fin(n)$ of naturals less than~$n$, given that these sets have a decidable equality and the arity of a type construct is finite.
Indeed, in our formalisation, we adopt $\Fin(n)$ as the set of type variables in the definition of $\Type_{\Sigma}$\ifarxiv (see \Cref{sec:formalisation} for details)\fi.
For the sake of clarity in presentation, though, we just use named variables and assume that $\Xi$ has a decidable equality.}
\end{remark}
We need some auxiliary definitions for the notion of extension to state the unification problem:
\begin{defn}
By an \emph{extension} $\sigma \geq \rho$ of a partial substitution $\rho$ we mean that the domain $\dom(\sigma)$ of $\sigma$ contains the domain of $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in $\dom(\rho)$.
  By a \emph{minimal extension} $\bar{\rho}$ of $\rho$ satisfying $P$ we mean an extension $\bar{\rho} \geq \rho$ with $P(\bar{\rho})$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma \geq \rho$ and $P(\sigma)$.
\end{defn}
\begin{lemma}\label{lem:unify}
  For any $A$ of $\Type_{\Sigma}(\Xi)$, $B$ of $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, either
  \begin{enumerate}
    \item there is a minimal extension $\bar{\rho}$ of $\rho$ such that $\simsub{A}{\bar{\rho}} = B$, or
    \item there is no extension $\sigma$ of $\rho$ such that $\simsub{A}{\sigma} = B$
  \end{enumerate}
\end{lemma}
This lemma can be inferred from the correctness of first-order unification~\citep{McBride2003,McBride2003a}, or be proved directly without unification.
We are now ready for \cref{thm:bidirectional-type-synthesis-checking}:

\begin{proof}[of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and independent of mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, so we only detail the last case \Rule{Op} which is new and has been discussed informally above.
  For brevity we omit the subscript $(\Sigma, \Omega)$.
